---
title: 'Project #1'
author: "YiWei Yang (18422148)"
date: '2018-03-12'
output: html_document
---
##Methods
```{r setup, include=FALSE}
library("tidyverse")
```

```{r}
library("phyloseq")
library("magrittr")
```

##Results:

###Data were loaded into R
```{r}
load("~/Documents/MICB425_portfolio/mothur_phyloseq.RData")

```



###Samples were rarefied/normalized to 100,000 sequences per sample to facilitate comparisons between samples. A random seed was set to ensure reproducibility.

```{r}
set.seed(4832)
m.norm = rarefy_even_depth(mothur, sample.size=100000)
```

###Rarefied counts were converted to relative abundance percentages.

```{r}
m.perc = transform_sample_counts(m.norm, function(x) 100 * x/sum(x))
```

##How does microbial community structure change with depth and oxygen concentration?

###Taxa presence and abundance
####Example plots using phyloseq for domain data. You should explore other taxonomy levels.
```{r}
m.perc %>% 
  
plot_bar(fill="Domain") + 
  geom_bar(aes(fill=Domain), stat="identity") +
  labs(title="Example 4: Domains across samples")

```

```{r}
m.perc %>% 
  
plot_bar() + 
  geom_bar(aes(fill=Domain), stat="identity") +
  facet_wrap(~Phylum, scales="free_y")+
  labs(title="Example 5: Phyla across samples")
```

####Examples outside of phyloseq

```{r}
m.perc %>%
  tax_glom(taxrank = 'Domain') %>%
  psmelt() %>% 

ggplot() +
  geom_boxplot(aes(x=Domain, y=Abundance)) +
  coord_flip() +
  labs(title="Example 6: Domain boxplots")
```

##Within your taxon, what is the richness (number of OTUs/ASVs)?

####Across all samples, there are 44 OTUs (i.e. taxa) within the unclassified domain.
```{r}
m.norm %>% 
  subset_taxa(Domain=="unknown") 
```

####Counting OTUs within the unclassified domain within each sample…

```{r}
m.norm %>% 
  subset_taxa(Domain=="unknown") %>%
  estimate_richness(measures = c("Observed"))
```

